---
title: "Example on how to use \"bigstatsr\""
# author: "Florian Privé"
date: "February 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center',
                      out.width = 600, out.height = 480)

require(bigsnpr)
celiac <- snp_attach("../../thesis-celiac/backingfiles/celiac_sub2_impute1.desc")
y <- celiac$fam$affection
pop <- celiac$fam$pop
```

## Introduction on the data

We attach a `big.matrix` from a backing file (stored on disk). It is a matrix of `char` (one byte signed integer) representing genotypes of a case-control study on [celiac disease](https://en.wikipedia.org/wiki/Coeliac_disease). 
```{r}
require(bigstatsr)
X <- attach.big.matrix("../../thesis-celiac/backingfiles/celiac_sub2_impute1.desc")
```
 
```{r}
dim(X)
typeof(X)
```

We also have, for each individual of the study, their disease status (1 for a control; 2 for a case) and their country(4 for the UK; 1 for the Netherlands; 2 for Italy; 5 for Finland).

```{r}
table(y)
rle(pop)
```

## Get more information on the data

We would like to know more about our data. For example, let's get the number of missing values for every column.

```{r}
parallel::detectCores() # number of available cores

system.time(
  nbNA <- big_apply(X, FUN = function(x) colSums(is.na(x)), 
                    .combine = 'c', block.size = 1000, ncores = 11)
)

all(nbNA == 0)
```

Here, we used the `big_apply` function which splits the data into column chunks of the `big.matrix`, load them in memory, apply a standard __R__ function and then combine the results. The `block.size` parameter controls the trade-off between speed and memory usage. Here we can see there is no missing value in our dataset (because it has already been imputed).

## Some statistical analysis

Let's apply Partial Singular Value Decomposition (PSVD) on our data to learn more about it.

```{r}
system.time(
  X.svd <- big_randomSVD(X, big_scale(center = TRUE, scale = TRUE),
                     k = 10, ncores = 11)
)
```

To get the scores of Principal Component Analysis (PCA), use 
```{r}
scores.PCA <- big_predScoresPCA(X.svd)
dim(scores.PCA)
plot(scores.PCA, col = pop, pch = 19, cex = 0.5)
```
We can see that PCA captures the "population structure" of the data. 

## Importance of each feature

We now want to analyze the disease status and firstly see which features (columns) are associated with the disease. 

```{r}
system.time(
  gwas <- big_univRegLog(X, y - 1, covar.train = scores.PCA, ncores2 = 6)
)
```

The previous function trains `r ncol(X)` models, i.e. one for each column of the `X`. Each model is a logistic regression including one column of `X`, an intercept and the matrix of covariables (here, the PCs to correct for possible confounders, such as population structure). 

```{r}
plot(-log10(gwas$p.value), pch = 19, cex = 0.5)
```

We can see that around the 190,000-th feature, there are some features that are highly correlated with the disease.

## A more sophisticated model

Let's now train a model on all the matrix at once, a logistic regression with elastic-net regularization.

```{r}
ind.train <- sort(sample(nrow(X), 10e3))
scores.PCA <- matrix(0, nrow(X), 10)
scores.PCA[] <- rnorm(length(scores.PCA))
system.time(
  log.mod <- big_spRegLog(X, y01.train = y[ind.train] - 1, ind.train = ind.train,
                          covar.train = scores.PCA[ind.train, ], 
                          penalty = "enet", alpha = 0.5, 
                          dfmax = 20e3, ncores = 11)
)
```
#### TODO: Use a predict function for models
