---
title: "Read a FBM from a file"
author: "Florian Privé"
date: "July 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this vignette, you learn how to read a Filebacked Big Matrix from a text file. [Package {bigreadr}](https://github.com/privefl/bigreadr) is required.

## Data

```{r}
library(bigstatsr)
library(bigreadr)

## LONG CSV
df <- datasets::mtcars
csv <- fwrite2(df[rep(seq_len(nrow(df)), 500000), ], 
               tempfile(fileext = ".csv"), 
               row.names = TRUE)
format(file.size(csv), big.mark = ",")
```

## Check file content

```{r}
nlines(csv)
(first_rows <- fread2(csv, nrows = 5))
sapply(first_rows, typeof)
ncol(first_rows)
```

What you can see with these first lines:

- there are some column names (header),
- each line has 12 elements,
- the first element of each line stores some row names,
- and the other elements seems to be all numeric.

## Read those data

- Read all numeric columns in an FBM
    ```{r}
    (test <- big_read(csv, select = 2:12))
    rbind(csv, test$backingfile)
    attr(test, "fbm_names")
    ```

- Read other non-numeric data afterwards
    ```{r}
    meta <- fread2(csv, select = 1)
    head(meta)
    ```

## Read filtered data

```{r}
## Get the filter data
filter <- fread2(csv, select = "cyl")[[1]] == 4
## Read only rows corresponding to 'filter'
(test2 <- big_read(csv, select = 2:12, filter = filter,
                   backingfile = tempfile()))
```
